\documentclass[12pt]{article}
\title{PyWIP User's Guide}
\author{Nicholas Chapman}
\date{9 July 2013}
\newcommand{\pywip}{PyWIP}
\begin{document}

\maketitle
\tableofcontents
\newpage

\section{Introduction}

Around the time I was supposed to be writing my thesis (way back in 2007), I
decided I was fed-up with a plotting package called WIP.  In Astronomy, people
generally use IDL, SM, WIP, or Gnuplot for their plotting needs.  The first two
cost money, a lot of money in the case of IDL.  WIP has two main advantages over
Gnuplot: it can read and understand FITS/MIRIAD data files with world
coordinates, and I know how to use it.  In most other ways WIP sucks.  Big time.

So, like I said, I was supposed to be writing my thesis, which was going to
require a ton of figures, and the thought of doing it all in WIP gave me
nightmares.  Right then and there, I decided it would be much easier to write
a python wrapper around WIP to make it much easier to use.  Thus, \pywip{} was
born.

The only other python plotting package I knew of at the time was matplotlib.  I
had played with that off and on, but was ultimately unsatisfied by it.  It
seemed to require too much overhead to make any sort of plot.  Furthermore, it
took major effort and more overhead to do some simple things like changing the
tick marks on the axes.  Lastly, matplotlib seemed to run very slowly, even on
the simplest of plots.  But, your mileage may vary.  If you hate \pywip{}, give
it a try. matplotlib can do many things that \pywip{} (and WIP) cannot. 
However, I do not need any of those features (usually).  There is also an
astronomy python plotting package based on matplotlib called APLpy.

\section{Installation}

\pywip{} is a python package, so all the code, documentation, and examples are
located inside a directory named \texttt{pywip}.  It is easy to install. You
only need two things, WIP and python.  You likely already have python
installed.  If you have a semi-recent version, greater than say, around 2.4,
that should be fine.  WIP comes from installing MIRIAD.

The most convenient way to install \pywip{} is to put it in some permanent
location and make that location your PYTHONPATH.  For this to work
you need to set the PYTHONPATH environment variable, e.g.:\\ \texttt{setenv
PYTHONPATH /Users/nchapman/data1/Applications/nlcpython}

\section{Documentation}

\pywip{} comes with numerous examples in the \texttt{bin} directory. You should
be able to run them all and get output postscript plots.  Furthermore, you can
find this users manual in \LaTeX{} and pdf format in the \texttt{docs}
directory.  Lastly, you can easily create an HTML guide that lists all the
commands with brief descriptions of their keywords:\\
\texttt{pydoc -w pywip}\\
\texttt{pydoc -w pywip.tools}

\section{Individual Commands}

This section will further describe individual commands.  It will usually 
describe `gotchas' and the like that may cause headaches.

\subsection{axis}

If you issue the \texttt{axis()} command without any arguments, then \pywip{}
will attempt to guess the appropriate axis to draw.  This means if you set
the \texttt{logx} or \texttt{logy} keywords in any of your plots, then the
appropriate x- or y-axis will be labeled logarithmically.  Furthermore,
if you plotted an image with \texttt{head=`rd'}, then \pywip{} will label
the axes with world coordinates.  Finally, when you are plotting multiple
panels that join, \pywip{} will be smart enough to suppress number labeling
where it would interfere with an adjacent panel.

This command isn't perfect, but it's good enough for Rock-and-Roll.  Give it a
whirl, and if it doesn't do it the way you like, then you can fall back on
specifying some or of all the parameters manually as needed.  Table
\ref{tab:axis} lists the allowed parameters.

\begin{table}
\caption{\label{tab:axis} Keywords for the \texttt{axis()} command}
\begin{center}
\begin{tabular}{lp{4in}}
\hline
\hline
\multicolumn{1}{c}{Keyword} & \multicolumn{1}{c}{Meaning}\\
\hline
box       & Specifies the frame to be drawn.  Defaults to
            \mbox{(`bottom',`left',`top',`right')}.\\
number    & Where to put the numbers.  Defaults to (`bottom',`left')
            but \texttt{axis()} will attempt to change these depending if you
            have multiple panels.  The allowed strings are: `left', `right', 
            `bottom', and `top'.  If you don't want any numbers, set this to ().\\
tickstyle & Style for tickmarks.  Note, to control whether tickmarks
            are actually shown, see the \texttt{subtick} and \texttt{majortick}
            keywords below.  The default style is (`inside',`inside'), meaning
            the tickmarks for (x,y) axes are drawn inside of the
            frame.  Either string can be set to `inside', `outside',
            or 'both'.\\
format    & Format for the numbers.  The default is (`auto',`auto')
            for the (x,y) axes.  If you are plotting an image, these
            are automatically changed to (`wcs',`wcs').  Generally, you should
            never have to set this, but the other options are `dec'
            to force decimal labeling, and `exp' to force exponential
            labeling.\\
xinterval & A tuple of two numbers, where the first number is the
            distance between major xtick intervals, and the second
            number is the number of subticks between each major xtick.
            If you leave this unset, reasonable defaults will be
            chosen.\\
yinterval & A tuple of two numbers, where the first number is the
            distance between major ytick intervals, and the second
            number is the number of subticks between each major ytick.
            If you leave this unset, reasonable defaults will be
            chosen.\\
\hline
\end{tabular}
\end{center}
\end{table}

\setcounter{table}{0} % need to have a two-page table

\begin{table}
\caption{Keywords for the \texttt{axis()} command (continued)}
\begin{center}
\begin{tabular}{p{1in}p{4in}}
\hline
\hline
\multicolumn{1}{c}{Keyword} & \multicolumn{1}{c}{Meaning}\\
\hline
drawtickx, drawticky   & If set to False, this is a shortcut to setting
                         \texttt{subtickx=False} and \texttt{majortickx=False}.
                         Note, setting \texttt{drawtick=False} while also
                         setting the \texttt{majortick} or \texttt{subtick}
                         keywords to True may cause weird behavior.  You have
                         been warned. \\
firstx, firsty         & If set to False, write only the last part of the label
                         for the first time tick on the axis.  For example, if
                         the full first label is 17 42 34.4, then only write
                         34.4. This option is useful for sub-panels that join
                         each other. The default values are True.\\
gridx, gridy           & Draw a grid of lines at the major tick positions for
                         either x or y.  Defaults are False, meaning no grid is
                         shown.\\
logx, logy             & Label x or y axis logarithmically.  Defaults are set
                         automatically depending on whether the user
                         has specified log labeling with plot() or some other
                         command. Note that setting this parameter will override
                         any settings for xinterval and yinterval.\\
subtickx, subticky     & Draw subticks?  Defaults to True, meaning do draw
                         subticks for both the x and y axes.\\
majortickx, majorticky & Draw majorticks? Defaults to True, meaning do draw
                         major ticks for both the x and y axes.\\
verticaly              & Rotate the numbers on the y axis to be vertical?  The
                         default is False, but it is set to True automatically
                         if an image has been plotted.  There is not equivalent
                         command for the xaxis numbers.\\
zerox, zeroy           & If set to False, omit leading zero in numbers $< 10$
                         in time labels.  This option is only valid when format
                         is `wcs' for the given axis.  Furthermore, it is
                         ignored when verticaly=True because it becomes
                         impossible to align the numbers properly.  The default
                         values are True.\\
\hline
\end{tabular}
\end{center}
\end{table}

\subsection{blowup}

This command is somewhat tricky.  The input coordinates MUST be pixel values. I
usually determine them from DS9.  Secondly,  if you have plotted only a
sub-region of an image with halftone/contour, then you need to input the
relative pixel values.  For example, I plotted the image
'image.fits[172:194,136:158]'.  Later, I wanted to draw a blow-up box determined
by the pixel coordinates 181.156,183.018,145.567,146.802 (in the xmin, xmax,
ymin, ymax format).  However, I could not enter these values for blowup()
because they are incorrect.  The correct values would be: 10.156, 12.018,
10.567, 11.802.  (181.156 - 171, 183.018 - 171, 145.567 - 135, 146.802 - 135).
I subtracted 171 and 135 for x and y respectively, because pixels in FITS images
are counted from 1 not zero.

\subsection{compass}

A kludgy \texttt{compass()} command has been added which should allow one to
plot the directions of galactic $\ell$ and $b$ on an RA/DEC plot.  It hasn't
been well-tested.

\subsection{legend}

The legend command is used to make it easier to put legends on your plots. A
legend is essentially a series of rows, where each row has two elements: a
symbol or line representing the data set or curve and a text label. After vast
amounts of work, I think I finally succeeded in making a general purpose
\texttt{legend()} command that doesn't require lots of fiddling by the user. WIP
does not come with such a command, so you should be awed by this achievement.
Just give it x and y coordinates for the upper-left corner (given as a fraction
of the plot width and height) and you should be good to go. Maybe someday I will
get around to adding the option to have a box drawn around the legend
automatically. The only caveat with \texttt{legend()} is that it should be the
last thing plotted (well, I think \texttt{xlabel},\texttt{ylabel}, and
\texttt{title()} are okay).

Also note, the curves listed by \texttt{legend()} are accumulated for all
panels as plots are made until a \texttt{legend()} command is issued.  Then
the list of plots is reset.  By default, plots are not shown in the legend
because they have \texttt{text=None} keyword.

\subsection{vector}

The \texttt{vector} command is used to draw arrow fields.  So that the angles
and lengths will always make sense when plotting arrows on a FITS image with
WCS, internally \texttt{vector()} will reset the header to `px' (absolute pixel
value).  What this means for the user is that \texttt{vector()} should be the
last thing added to a plot (well, maybe before \texttt{legend()}, I'm not sure
about that).  Also, it means that you should always specify the x and y as
pixels, not wcs coordinates.  Otherwise, your positions, lengths, and angles will
be all wrong.  Finally, you can use the \texttt{align} keyword to  specify
whether the input coordinates are the left (end), center, or right (tip) of the
arrows.

You can also plot polarization `sticks' (vectors without arrowheads) by 
setting \texttt{taper=0}.

\section{Keywords}

\subsection{color,font,size,style,width}

A number of the commands have some or all of these keywords as options.  If
some or all of these aren't specified, the default values for these will be
used.  The default color is black, font is roman, size is 1, style is
solid lines, and width is 1.  You can change the defaults at any time by
running the \texttt{default} command.  See Tables \ref{tab:colors},
\ref{tab:fonts}, and \ref{tab:lstyles} for details.

RGB colors can be used instead of the built-in defaults.  This hasn't been
fully tested and debugged, but seems to work well.  To specify an rgb color,
do it like so: \texttt{color='0,0.5,0'} which is 50\% green, and no red or
blue.

\begin{table}
\caption{\label{tab:colors} Valid Colors}
\begin{center}
\begin{tabular}{cl}
\hline
\hline
\multicolumn{1}{c}{Key} & \multicolumn{1}{c}{Description}\\
\hline
w  & white\\
k  & black (default)\\
r  & red\\
g  & green\\
b  & blue\\
c  & cyan\\
m  & magenta\\
y  & yellow\\
o  & orange\\
gy & green-yellow\\
gc & green-cyan\\
bc & blue-cyan\\
bm & blue-magenta\\
rm & red-magenta\\
dg & dark gray (approx gray65)\\
lg & light gray (aprox gray35)\\
gray1-100 & Various shades of grayscale (1 is white, 100 is black)\\
r,g,b & rgb color as fractions from 0-1 \\
\hline
\end{tabular}
\end{center}
\end{table}

\begin{table}
\caption{\label{tab:fonts} Valid Fonts}
\begin{center}
\begin{tabular}{cl}
\hline
\hline
\multicolumn{1}{c}{Key} & \multicolumn{1}{c}{Description}\\
\hline
sf & sans-serif\\
rm & roman (default)\\
it & italics\\
cu & cursive\\
\hline
\end{tabular}
\end{center}
\end{table}

\begin{table}
\caption{\label{tab:lstyles} Line Styles}
\begin{center}
\begin{tabular}{cl}
\hline
\hline
\multicolumn{1}{c}{Key} & \multicolumn{1}{c}{Description}\\
\hline
  -  & solid (default)\\
 \texttt{--}  & dashed\\
 .-  & dot-dashed\\
  :  & dotted\\
-... & dash-dot-dot-dot\\
\hline
\end{tabular}
\end{center}
\end{table}

\subsection{image}

The \texttt{contour}, \texttt{halftone}, and \texttt{winadj} commands all have
the \texttt{image} keyword.  \texttt{image} is a string specifying the
FITS/MIRIAD image name, and optionally the sub-image to plot.  For example:\\
\texttt{image='Bob.fits'} Plot the entire image `Bob.fits'\\
\texttt{image='Bob.fits[10:20,5:15]'} Plot `Bob.fits' but only the subimage of
pixels 10-20 in the x and 5-15 in the y.  Finally, \texttt{image='Bob.fits(1)'}
will plot plane 1 from `Bob.fits'.  You can combine the subimage and plane
definitions in either order.

There are several other `gotchas' to watch out for when plotting images.
See \S\,\ref{sec:futz-images} for more details on these.

\subsection{limits}

The limits keyword can be have several different values.  First, if you do not
specify limits, then \pywip{} will attempt to guess what the limits should be.
If no limits have been set for the current panel, then \pywip{} will use the 
min/max of the dataset for the current limits.  Otherwise, the currently
existing limits will be reused.  Second, you can give a list/tuple of four
values, which will become the new limits and override any previously existing
ones.  Third, you can set \texttt{limits='last'}, meaning \pywip{} should
use the last known limits (generally from the previous panel) as the new limits.
Note, this will also carry over settings for \texttt{'logx'} and 
\texttt{'logy'}.

The fourth, and last possibility is to explicitly set \texttt{'limits=None'}. I
haven't fully explored the ramifications of this, but it seems to force \pywip{}
to set new limits based on the min/max of the dataset or image, even if limits
were already known for the current panel. I use it in \texttt{example11.py} to
ensure that the blowup box is not resized to the limits of the entire region. I
think it would work for non-image type plots as well.

\subsection{palette}

The \texttt{halftone()} command allows a palette to be a filename which contains
an RGB lookup table.  I had some grandiose plans when I first implemented this,
but in reality it is quite limited.  I think PGPLOT has a limit of 255 color
levels, so you cannot make a large-scale RGB plot.  This lookup file needs to
have four columns given as red,green,blue,scol.  The red, green, and blue 
values must be given as fractions between 0 and 1.  The same is true for scol,
which defines fractions of the maximum intensity.  Linear interpolation for
values in the image between specified levels will be performed.

To make a proper RGB plot, you would need to create a fake image (no larger
than $16\times16$, and give each pixel an increasing intensity.  Then, write
a lookup table and for each intensity specify what rgb color it should have.
You can see why this is a pain in the ass, and shouldn't be done.  Easier to
just use DS9.

\subsection{x,y}

Several commands have the \texttt{x} and \texttt{y} keywords to specify a
location on the plot.  This can be given as a number, but it may be more
convenient to specify a string in the case of world coordinates.  For example,
\texttt{x='3:30'} would move the cursor to 3 hours 30 minutes.  Similarly,
\texttt{x='52.5'} would also move the cursor to 3 hours 30 minutes since 52.5
degrees is the equivalent.  Note the importance of giving the value as a string.
Without the quotes, it will move the cursor to 52.5 hour-seconds, not 52.5
degrees, nor 3 hours 30 minutes.  \pywip{} assumes that \texttt{x} is for Right
Ascension and \texttt{y} is for Declination.

\section{Other Tools}

\pywip{} includes two additional commands located in the tools subpackage.
They are \texttt{imextract} and \texttt{levelgen}.  These tools require
\texttt{numpy} and \texttt{pyfits} to use.

Furthermore, there are several stand-alone programs in the \texttt{pywip/bin}
directory that may be useful.  All of them use \texttt{readcmd.py} to read
command line arguments.  These are described below.

\subsection{ds9towip}

This program will read contour files created by DS9 and convert them to
ones readable by WIP.  It will create an output directory, and then inside
that a directory a series of files, one for each contour level.  Lastly, it
will create a WIP file that can be processed using \texttt{inputwip()}.  This
script is useful because DS9's contouring algorithm is superior to WIP's in
some circumstances (avoiding null pixels at the edges of maps).

\subsection{qplot}

\texttt{qplot.py} is a quick plotting script that can read a file from the
command line and implements many common plotting commands.  The idea is that if
you just want to do a simple plot of your data, you could use \texttt{qplot.py}
instead of writing a python script.

\subsection{rotatevector}

Because of projection effects, north on your map may not point exactly vertical.
However, polarization vectors are usually defined as an angle east of north.
Thus, \texttt{rotatevector.py} will rotate the vectors so that their angles
will correspond to degrees east of vertical.  This should never change your
vector angles by more than a degree or two at most, so this is a really minor
effect.

\subsection{tickmarks}

One of the many limitations of WIP is that it doesn't properly handle world
coordinates for large regions on the sky.  To fix this, I wrote
\texttt{tickmarks.py}. You can use it to generate a file of raw WIP commands
which can be used in your script with the \texttt{inputwip()} command.
\texttt{tickmarks.py} manually draws the tickmarks and labels, and does a pretty
good job of guessing things.  The only drawback is that you have to manually
add the label for the first tickmark along the x axis by editing the output
WIP file created by \texttt{tickmarks.py}.  It is pretty easy to do.  Just
scroll down to where the RA labels are defined.

\subsection{worldpos}

When I first wrote this, I was ecstatic.  At the time it was the only pure
python implementation of pixel $\leftrightarrow$ world coordinate conversion. 
You could still \hfill use \hfill it \hfill for \hfill that \hfill purpose
\hfill if \hfill you \hfill like, \hfill and \hfill in \hfill fact \hfill 
\texttt{tickmarks.py} \hfill and  \\
\texttt{rotatevector.py} do so.  However, there is
now an ``official'' python package, \texttt{pywcs} that does the same thing and
is more sophisticated about handling rotations.

\section{Tips \& Tricks}

\subsection{Using hidden variables}

Say you want to make a series of plots in a for loop, and have each plot be a
different color.  You don't need to manually type in all the colors.  Instead,
they are part of a hidden variable that can be imported.  Hidden variables
the user may want to access are listed in Table \ref{tab:hidden}.  They
can be imported and used like so:\\
\texttt{from pywip import \_colors as pycolors}

\begin{table}
\caption{\label{tab:hidden} Hidden Variables}
\begin{center}
\begin{tabular}{ll}
\hline
\hline
\multicolumn{1}{c}{Variable} & \multicolumn{1}{c}{Meaning}\\
\hline
\_palettes & The palettes used for the \texttt{halftone} command\\
\_colors   & The valid colors for fonts, symbols, lines, etc. (see Table
             \ref{tab:colors})\\
\_fills    & The available fill styles\\
\_fonts    & The available fonts (See Table \ref{tab:fonts})\\
\_lstyles  & The available line styles (Table \ref{tab:lstyles})\\
\_symbols  & The available symbols to plot points\\
\hline
\end{tabular}
\end{center}
\end{table}

\subsection{Multi-Variable Axes}

To plot one variable (with numbers and tickmarks) on the bottom and a second
variable  on the top (with a different set of numbers and tickmarks), you
have to use the \texttt{style=None} keyword in the \texttt{plot()} command.
This generates a phantom plot.  \pywip{} goes through the motions of setting
limits, but won't actually plot any points.  Then, just plot the new axis:
\texttt{axis(box=['top'],number=['top'])}.  You will also have to change the 
axis command for the first plot so that it doesn't interfere with the second
\texttt{axis()} command, e.g.\ \texttt{axis(box=['bottom','left','right'])}.

I did this, and it seems like it would work, so long as variable two is 
directly proportional to variable1.  For example, $var2 = 3\times var1$ should
be okay.  However, I had a case where $var2 = 1/var1$, and that did NOT work.
I think it is a problem with WIP however.

\subsection{\label{sec:futz-images}Futzing with images}

WIP is particular about the images it will accept. Images in \pywip{} will
always have x increasing to the right and y increasing upwards.  If you are
plotting an image with a world coordinate system (WCS), it is up to you to make
sure the image is rotated properly.  In astronomy you typically want Declination
upwards and Right Ascension increasing to the left.  You may need to rotate your
image with IRAF's \texttt{rotate} command or the \texttt{regrid} task in Miriad
to achieve the correct rotation.  If you neglect to do this, \pywip{} will still
do something, but the WCS will probably be wrong.

Secondly, at least one world coordinate projection (gls) is not allowed.  If
you get an error about ``unsupported projection'' then you may need to fix your
image.  This can be done with the Miriad task \texttt{regrid}.  Thanks to 
Jorge Pineda for this tip:
\begin{verbatim}
Assume you have file.fits
fits in=file.fits op=xyin out=file.mir
regrid in=file.mir out=file_tan.mir project=tan
fits in=file_tan.mir op=xyout out=file_tan.fits
\end{verbatim}

Lastly, for very large fields, I think the WCS displayed by WIP will be off. 
This can easily be seen if you plot the ra/dec, then plot their equivalent x/y
positions on an image (get the x/y through something like \texttt{sky2xy} or
\texttt{ds9}). They will \emph{not} line up.

\subsection{World Coordinates from a data file}

When reading from files, WIP needs ra/dec to be in hour-seconds and 
degree-seconds (ra * 240, when ra is in degrees, and dec * 3600 when dec is
in degrees).

\section{The Future}

\pywip{} is starting to show its age.  It is based on PGPLOT and WIP, which
means many of the underlying problems cannot be fixed.  Since I first started
\pywip, I have learned \texttt{numpy} and \texttt{pyfits}, and use them both
extensively, which eliminates any of the speed advantage of \pywip{} over
matplotlib.  But, I still think the syntax of \pywip{} rules, so I am learning
matplotlib and implementing wrappers around their commands to make them easier
to use.  All in all, the need for \pywip{} is less than it used to be. As a
result, this is the last version of \pywip.  It is possible I made make some bug
fixes in the future, but there are no guarantees

For historical reasons, here is my final list of \pywip{} limitations:

\begin{itemize}
\item The colors should allow hexcodes, grayscale ranges, and rgb values
(partially fixed)
\item The axis() command should control logx, logy, and plot range limits
\item Cannot make shaded contour plots
\item Cannot make translucent fill styles (I don't think PGPLOT allows
transparency)
\item Only a limited number of filled symbols are available
\item The only output formats are .ps and .gif. (PGPLOT problem again)
\item The legend() command is quite nice, but is still a kludge
\item No ability to make RGB plots like DS9 (limited to 255 levels)
\item WCS projections do not seem correct for large regions
\item No way to vertically align text (in a left, center, right format)
\item Doesn't understand numpy arrays
\item you should be able to specify whether a given set of coordinates are
viewport, wcs, or whatever.
\item filled histograms
\item an easy scalebar (i.e.\ draw a line and specify distance in pc, whatever)
\end{itemize}

\end{document}
